---
title: "R Workshop"
subtitle: "Plotting, data import, and data processing"
author: "Gavin Simpson"
date: "2018-06-15"
output:
  xaringan::moon_reader:
    css: ['default', './resources/my.css']
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA,
                      echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE,
                      fig.align = 'center', fig.height = 4.5, fig.width = 8, dev = 'svg')
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
options(htmltools.dir.version = FALSE)
```
```{r packages, include = FALSE, cache = FALSE, echo = FALSE}
library('ggplot2')
library('cowplot')
library('tidyr')
library('knitr')
library('viridis')
library('dplyr')
library('readr')
library('readxl')
```
```{r ggplot-set-up, cache=FALSE, echo = FALSE}
theme_set(theme_bw())
```

# Preamble

Install some packages

```{r package-install, eval = FALSE}
pkgs <- c('ggplot2', 'cowplot', 'tidyr', 'viridis', 'dplyr', 'readr',
          'readxl', 'gapminder')
install.packages(pkgs)
```

Code / slides are available:

[https://github.com/gavinsimpson/ipa-2018-rstats]()

---

# Warm-up

```{r intro-figure, echo = FALSE}
## install.pckages('gapminder')
data(gapminder, package = 'gapminder')

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 1, aes(color = continent)) +
  stat_smooth(method = "lm", size = 1, color = "black") +
  scale_x_log10() +
  xlab("Per Capita GDP") +
  ylab("Life Expectancy (yrs)") +
  facet_wrap(~continent, scales = "free") +
  theme_bw() +
  guides(color=FALSE)
```
---

# Intro

```{r intro-figure-code, eval = FALSE, ref.label = 'intro-figure', dependson = -1}
```

---

# Plotting in R

Three main ways to plot in R

1. *base graphics* &mdash; what most of R's out-of-the-box `plot` code uses,
2. *lattice* &mdash; an R implementation of Trellis graphics in S-PLUS
3. *ggplot2* &mdash; an R implementation of the Grammar of Graphics
---

# Why choose *ggplot2*?

* Compact & more expressive code than base graphics
* Plots look nicer (to me at least) than either base graphics or lattice
* Automates production of legends and keys
* Powerful for exploratory data anaylsis & quickly interating on plots
* Widely-used, popular &mdash; lots of help/guides available on-line

---

# The grammar

* data
* aesthetics
* **geom**etry
* stats
* facets
* scales
* coordinates

---

# Data

* The data object (or objects) for a plot must be in the form of a *tidy* data frame
* Can use one data frame for all elements of a plot
* But individual layers can have their own data

---

# Edgar Anderson's `iris` dataset

```{r head-iris}
head(iris)
```

* These data are stored in a wide format

* Information is stored in the column labels

* One column per variable

---

# Tidy data

* Tidy data is a concept of Hadley Wickham's, author of *ggplot2*

* Similar to long format or normalised data

* If you want to assign a variable to an element on the plot it needs to be a variable in the tidy data frame

---

# `tidyr`

```{r}
iris[1:2, ]

library('tidyr')
df <- gather(iris, key = "attribute", value = "measurement",
             - Species)

df[1:2, ]
```

---

# Climate data &mdash; reading in data

In-built reading of tabular data from files

```{r}
climate <- read.csv('http://bit.ly/2JLBIFv')
climate <- climate[, -1]
climate[1:2, ]
climate[1:2, ]
```

---

# Reading in data via *tidyverse*

Consistent versions from the *tidyverse*

```{r}
library('readr')
climate <- read_csv('http://bit.ly/2JLBIFv')
climate <- climate[, -1]
climate[1:2, ]
```

